---
title: "Phenotypic change"
author: "Julius Hoffmann"
date: "2024-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r , message=FALSE}
library(tidyverse)
library(gridExtra)
library(data.table)
library(ggh4x)
library(flextable)
library(ggrepel)
```


```{r}
set_flextable_defaults(
  font.size = 10, theme_fun = theme_vanilla,
  padding = 6,
  font.family = "Arial", digits = 3, line_spacing = 1.3)
```


## Load data

```{r pressure, echo=FALSE}
OD_96h<- read.csv("../derived data/Clone_OD_96h.txt")
```

1) remove defence values > 0 from the data

```{r}
OD_96h[,c("gPred_gAlone" )] <-  replace(OD_96h[,c("gPred_gAlone" )], OD_96h[,c("gPred_gAlone" )] > 0, c(NA))
```

2) calculate median traits and corresponding quantiles

```{r}

Obs_traits_stats <- OD_96h %>% 
  group_by(prey,predator, replicate, experiment_day) %>% 
  reframe(median_df =  median(gPred_gAlone, na.rm = TRUE),
          median_growth = median(cloneOD, na.rm = TRUE),
          median_df_quant_1 = quantile(gPred_gAlone, probs = 0.25, na.rm = TRUE), 
          median_df_quant_2 = quantile(gPred_gAlone, probs = 0.75, na.rm = TRUE),
          median_growth_quant_1 = quantile(cloneOD, probs = 0.25, na.rm = TRUE), 
          median_growth_quant_2 = quantile(cloneOD, probs = 0.75, na.rm = TRUE))

```

## Statistical Analysis
1) Testing statistical significance of difference between carrying capacity and defence distributions on day 8 and day 60.

For defence:
```{r}
tests_defence <- OD_96h %>% 
  mutate(replicate = str_replace_all(replicate, c("1" = "A", "3" = "B", "5" = "C"))) %>% 
  filter(experiment_day %in% c(8,60)) %>% 
  group_by(prey,predator,replicate) %>% 
  reframe(p.value = kruskal.test(gPred_gAlone ~ experiment_day)$p.value,
          rank_sum_statistic = kruskal.test(gPred_gAlone ~ experiment_day)$statistic,
          Df = kruskal.test(gPred_gAlone ~ experiment_day)$parameter) %>% 
  ungroup() %>% 
  mutate(p.value_corr = p.adjust(p.value,method = "BH"))


ft_tests_defence <- flextable(tests_defence) 
ft_tests_defence <- colformat_double(x = ft_tests_defence, digits = 3)

ft_tests_defence 
save_as_image(ft_tests_defence,"../Results/kruskal_test_defence.png")
```

For carrying capacity:
```{r}
tests_carrying_cap <- OD_96h %>% 
  mutate(replicate = str_replace_all(replicate, c("1" = "A", "3" = "B", "5" = "C"))) %>% 
  filter(experiment_day %in% c(8,60)) %>% 
  group_by(prey,predator,replicate) %>% 
  reframe(p.value = kruskal.test(cloneOD ~ experiment_day)$p.value,
          rank_sum_statistic = kruskal.test(cloneOD ~ experiment_day)$statistic,
          Df = kruskal.test(cloneOD ~ experiment_day)$parameter) %>% 
  ungroup() %>% 
  mutate(p.value_corr = p.adjust(p.value,method = "BH")) 


ft_tests_carrying_cap <-flextable(tests_carrying_cap) %>% 
                                colformat_double(digits = 3)
ft_tests_carrying_cap

save_as_image(ft_tests_carrying_cap,"../Results/kruskal_carrying_cap.png")
```


2) Testing correlation between carrying capacity and defence based on median values per microcosm and sampling day.

```{r}

median_traits_corr <- OD_96h %>% 
  group_by(prey,predator, replicate, experiment_day) %>% 
  reframe(median_df =  median(gPred_gAlone, na.rm = TRUE),
          median_growth = median(cloneOD, na.rm = TRUE)) %>% 
  ungroup() %>% 
  reframe(correlation = cor.test(median_growth, median_df, na.action("na.omit"),method = "kendall")$estimate,
          p_value = cor.test(median_growth, median_df, na.action("na.omit"),method = "kendall")$p.value) %>% 
  ungroup() %>% 
  mutate(p_value_corr = p.adjust(p_value,method = "BH"))%>% 
  flextable() %>% 
  colformat_double(digits = 4)
median_traits_corr

save_as_image(median_traits_corr,"../Results/correlation_defence_carrying_cap.png")
```


## Visualise Community Level Phenotypic Change

```{r, echo=FALSE}
prey.labs <- c("ancestral prey", "evolved prey")
names(prey.labs) <- c("Ancestral Bacteria", "Evolved Bacteria")

predator.labs <-  c("ancestral ciliates", "evolved ciliates")
names(predator.labs) <- c("Ancestral Ciliates", "Evolved Ciliates")

backgrounds <- list(NULL,element_blank())
texts<- list(NULL,element_blank())
disabled_strip <- strip_nested(
  background_x = backgrounds,
  text_x = texts,
  by_layer_x = TRUE)


Obs_traits_stats$day <- Obs_traits_stats$experiment_day
Obs_traits_stats$day[Obs_traits_stats$day == 28] <- ""

ggplot(data = Obs_traits_stats) +
  geom_point( mapping = aes(x = median_df, y = median_growth, shape = interaction(predator,experiment_day), colour = interaction(prey,replicate), fill=interaction(prey,replicate) ),size = 4.05, stroke = 1.6)+
    geom_label_repel( aes(x = median_df, y = median_growth, label = day),
                  box.padding   = 0.4, 
                  point.padding = 0.4,
                  segment.color = '#566573',
                  min.segment.length = 0.1,
                  label.size = 0.4,
                  label.padding = 0.15) +
  geom_path(mapping =aes(x = median_df, y = median_growth), colour = "#000000", linewidth = 0.8, alpha = 0.7) +
  geom_errorbar(mapping =aes(x = median_df, y = median_growth, ymin = median_growth_quant_1, ymax = median_growth_quant_2), width=.00, linewidth = 0.6, alpha = 0.6,colour = "#696969")+
  geom_errorbarh(mapping =aes( y = median_growth, xmin = median_df_quant_1, xmax = median_df_quant_2), height=.00, linewidth = 0.6, alpha = 0.7,colour = "#696969")+
  facet_nested(predator ~ prey +replicate , labeller = labeller(prey = prey.labs, predator = predator.labs), strip = disabled_strip)+
  
  
  scale_shape_manual("Median on day\n of experiment", breaks = c("Ancestral Ciliates.8","Ancestral Ciliates.28","Ancestral Ciliates.60","Evolved Ciliates.8", "Evolved Ciliates.28","Evolved Ciliates.60"), values = c(19,19,19,17,17,17) ,guide = "none")+
  scale_colour_manual(breaks = c("Ancestral Bacteria.1","Evolved Bacteria.1","Ancestral Bacteria.3","Evolved Bacteria.3","Ancestral Bacteria.5","Evolved Bacteria.5"), values = c("#7ead05","#c67cff","#7ead05", "#00bec4","#7ead05","#f8776e"),guide = "none")+
  scale_fill_manual(breaks = c("Ancestral Bacteria.1","Evolved Bacteria.1","Ancestral Bacteria.3","Evolved Bacteria.3","Ancestral Bacteria.5","Evolved Bacteria.5"), values = c("#7ead05","#c67cff","#7ead05", "#00bec4","#7ead05","#f8776e"),guide = "none")+
  
  labs( x = "defence [log( growth with predator / growth alone )]", y = "carrying capacity [optical density]")+
  xlim(-2.8,0)+
  ylim(0,0.75)+
  theme_bw()+
  theme( plot.title = element_text(size = 20, face = "bold"),
         legend.text = element_text(size = 18),
         legend.title = element_text(size = 18, face = "bold"),
         legend.key.size = unit(1, "cm"),
         axis.text = element_text(color = "grey50", size = 18),
         axis.title = element_text(size = 20),
         axis.title.y = element_text( colour = "black"),
         strip.text = element_text(size = 18, colour = "black", face = "bold"),
         strip.background = element_rect(color="black", fill="white", linewidth=1, linetype="solid"),
         
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), 
         axis.line = element_line(colour = "black"))  

#ggsave("Phenotypic_change_d8_d60.jpg", device = "jpg", path = "../Results", width = 450, height = 200,  units = "mm" )
ggsave("Phenotypic_change_d8_d60.pdf", device = "pdf", path = "../Results", width = 450, height = 200,  units = "mm" )

#, arrow = arrow(length=unit(0.35,"cm"),type = "closed",angle = 10)
#  scale_shape_manual("Median on day\n of experiment", breaks = c("Ancestral Ciliates.8","Ancestral Ciliates.28","Ancestral Ciliates.60","Evolved Ciliates.8", "Evolved Ciliates.28","Evolved Ciliates.60"), values = c(1,21,19,2,24,17) ,guide = "none")
#scale_fill_manual(breaks = c("Ancestral Bacteria.1","Evolved Bacteria.1","Ancestral Bacteria.3","Evolved Bacteria.3","Ancestral Bacteria.5","Evolved Bacteria.5"), values = c("#CAFF70","#FFBBFF","#CAFF70", "#97FFFF","#CAFF70","#FFA500"),guide = "none")
```

## Visualise overall trait distribution

1) Defence

```{r, echo=FALSE, fig.dim=c(10,8)}

OD_96h %>% 
 
ggplot()+
  geom_histogram(aes(x= gPred_gAlone))+
  facet_grid(predator ~ prey)+
  labs(x = "defence [log( growth with predator / growth alone )]")+
  theme_bw()+
  theme( plot.title = element_text(size = 20, face = "bold"),
         legend.text = element_text(size = 18),
         legend.title = element_text(size = 18, face = "bold"),
         legend.key.size = unit(1, "cm"),
         axis.text = element_text(color = "grey50", size = 18),
         axis.title = element_text(size = 20),
         axis.title.y = element_text( colour = "black"),
         strip.text = element_text(size = 18, colour = "black", face = "bold"),
         strip.background = element_rect(color="black", fill="white", linewidth=1, linetype="solid"),
         
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), 
         axis.line = element_line(colour = "black"))  

ggsave("Phenotypic_defence_distribution.jpg", device = "jpg", path = "../Results", width = 360, height = 200,  units = "mm" )

```

1) Carrying capacity

```{r, echo=FALSE, fig.dim=c(10,8)}

OD_96h %>% 

ggplot()+
  geom_histogram(aes(x= cloneOD))+
  facet_grid(predator ~ prey)+
  labs(x = "growth capacity [optical density]")+
  theme_bw()+
  theme( plot.title = element_text(size = 20, face = "bold"),
         legend.text = element_text(size = 18),
         legend.title = element_text(size = 18, face = "bold"),
         legend.key.size = unit(1, "cm"),
         axis.text = element_text(color = "grey50", size = 18),
         axis.title = element_text(size = 20),
         axis.title.y = element_text( colour = "black"),
         strip.text = element_text(size = 18, colour = "black", face = "bold"),
         strip.background = element_rect(color="black", fill="white", linewidth=1, linetype="solid"),
         
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), 
         axis.line = element_line(colour = "black"))  

ggsave("Phenotypic_growth_distribution.jpg", device = "jpg", path = "../Results", width = 360, height = 200,  units = "mm" )

```
