---
title: "Population Densities"
author: "Julius Hoffmann"
date: "2024-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load packages


```{r, message=FALSE}
library(tidyverse)
library(patchwork)
library(multcomp)
library(flextable)

```

```{r}
set_flextable_defaults(
  font.size = 10, theme_fun = theme_vanilla,
  padding = 6,
  font.family = "Arial", digits = 4)
```

## Load data


```{r, message=FALSE }
predator_density <- read_delim("../derived data/predator_densities.tsv")
prey_density <-read_delim("../derived data/prey_densities.tsv")

```

## Prepare data

Setting matching and informative variable names.

```{r, echo=FALSE}

#extract predator densities of interest and re-name cell content 
predator_density<- predator_density %>% 
  mutate(prey_history = str_replace_all(prey_history, c("ANC" = "ancestral prey", "EVO" = "evolved prey")),
         predator_history = str_replace_all(predator_history, c("ANC" = "naive predator", "EVO" = "evolved predator")))


#extract prey densities of interest and re-name cell content 
prey_density<-  prey_density %>% 
  mutate(prey_history = str_replace_all(prey_history, c("anc" = "ancestral prey", "evo" = "evolved prey")),
         predator_history = str_replace_all(predator_history, c("anc" = "naive predator", "evo" = "evolved predator", "none" = "no predator")))

#merge predator and prey dynamics
pred_prey_density <- right_join(predator_density,prey_density[,2:7])
colnames(pred_prey_density)[6] <- c("predator_mL")

```

## Calculating mean population densities

```{r}
pop_den_mean <- pred_prey_density %>% 
  group_by( predator_history,prey_history,replicate) %>% 
  reframe(mean_Pred= mean(predator_mL,na.rm = TRUE), #calculate predator mean within interval
          mean_OD= mean(bacteria_od,na.rm = TRUE)) #calculate prey mean within interval
```

## Calculate group means and standard error

```{r}
group_densities<-pop_den_mean %>% 
  group_by(predator_history,prey_history) %>% 
  reframe(avg_Pred= mean(mean_Pred), 
          se_Pred= sd(mean_Pred)/sqrt(length(mean_Pred)),
          avg_OD= mean(mean_OD),
          se_OD= sd(mean_OD)/sqrt(length(mean_OD)))
```


## Fitting a linear model

1) add distinguishable replicate names

```{r}
pop_den_mean <- pop_den_mean %>% 
  group_by(predator_history, prey_history, replicate ) %>% 
  mutate(Replicate_unique = cur_group_id() ) %>% 
  ungroup()
```

2) setting correct variable classes

```{r}
pop_den_mean$Replicate_unique <- as.factor(as.matrix(pop_den_mean$Replicate_unique))
pop_den_mean$predator_history <- as.factor(pop_den_mean$predator_history)
pop_den_mean$prey_history <- as.factor(pop_den_mean$prey_history)
pop_den_mean$replicate <- as.factor(pop_den_mean$replicate)

#Using ordered predictors
pop_den_mean$predator_history <- factor(pop_den_mean$predator_history, levels = c("no predator", "naive predator", "evolved predator" ))

pop_den_mean$prey_history <- factor(pop_den_mean$prey_history, levels = c("ancestral prey", "evolved prey" ))

```

3) running a linear model via glm() 

For predator densities:

```{r}
mod_den_Pred <- glm(mean_Pred ~ predator_history * prey_history ,family = gaussian(link = "identity"), data = pop_den_mean)
```

Model assessment predators:

```{r}
plot(mod_den_Pred)
plot(pop_den_mean$mean_Pred[1:12],fitted(mod_den_Pred))
abline(0,1)
```

Model inference for predators:

```{r}
#model output
tf_mod_den_Pred <- as_flextable(mod_den_Pred)
tf_mod_den_Pred <- set_caption(tf_mod_den_Pred, "GLM - predator densities", 
  style = "Table Caption") %>% 
                                colformat_double(digits = 4)

tf_mod_den_Pred

save_as_image(tf_mod_den_Pred,"../Results/GLM_predator_densiteis.png")
#ANOVA
mod_anova_Pred <- anova(mod_den_Pred, test = "F")
tf_mod_anova_Pred <- cbind(row.names(mod_anova_Pred[,1:6]), mod_anova_Pred[,1:6])
colnames(tf_mod_anova_Pred)[1] <- ("variable")
tf_mod_anova_Pred <- flextable(tf_mod_anova_Pred)%>% 
                                colformat_double(digits = 4)
tf_mod_anova_Pred <- set_caption(tf_mod_anova_Pred, "ANOVA - predator densities", 
  style = "Table Caption")

tf_mod_anova_Pred

save_as_image(tf_mod_anova_Pred,"../Results/ANOVA_predator_densiteis.png")
#pair-wise comparisons
pair_wise_Pred <-TukeyHSD(aov(mod_den_Pred))
ft_tukey_pred <- flextable(data.frame(variable = rownames(pair_wise_Pred$`predator_history:prey_history`),pair_wise_Pred$`predator_history:prey_history` ))%>% 
                                colformat_double(digits = 4)
ft_tukey_pred <- set_caption(ft_tukey_pred, "Pair-wise comparisons - predator densities", 
  style = "Table Caption")

ft_tukey_pred 

save_as_image(ft_tukey_pred ,"../Results/pair-wise_predator_densiteis.png")

#letters for multiple comparisons
multcompView::multcompLetters4(aov(mod_den_Pred), pair_wise_Pred)
```


For prey densities (OD):

```{r}
mod_den_OD <- glm(mean_OD ~ predator_history * prey_history ,family = gaussian(link = "identity"), data = pop_den_mean)
```
Model assessment prey:

```{r}
plot(mod_den_OD)
plot(pop_den_mean$mean_OD[1:18],fitted(mod_den_OD))
abline(0,1)
```


Model inference for prey:

```{r}
#model output
tf_mod_den_OD <- as_flextable(mod_den_OD)
tf_mod_den_OD <- set_caption(tf_mod_den_OD, "GLM - prey densities", 
  style = "Table Caption")%>% 
                                colformat_double(digits = 4)

tf_mod_den_OD

save_as_image(tf_mod_den_OD,"../Results/GLM_Prey_densiteis.png")
#ANOVA
mod_anova_OD <- anova(mod_den_OD, test = "F")
tf_mod_anova_OD <- cbind(row.names(mod_anova_OD[,1:6]), mod_anova_OD[,1:6])
colnames(tf_mod_anova_OD)[1] <- ("variable")
tf_mod_anova_OD <- flextable(tf_mod_anova_OD) %>% 
                                colformat_double(digits = 4)
tf_mod_anova_OD<- set_caption(tf_mod_anova_OD, "ANOVA - prey densities", 
  style = "Table Caption")


tf_mod_anova_OD

save_as_image(tf_mod_anova_OD,"../Results/ANOVA_prey_densiteis.png")
#pair-wise comparisons
pair_wise_OD <-TukeyHSD(aov(mod_den_OD))
ft_tukey_OD <- flextable(data.frame(variable = rownames(pair_wise_OD$`predator_history:prey_history`),pair_wise_OD$`predator_history:prey_history` ))%>% 
                                colformat_double(digits = 4)
ft_tukey_OD <- set_caption(ft_tukey_OD, "Pair-wise comparisons - prey densities", 
  style = "Table Caption")

ft_tukey_OD 

save_as_image(ft_tukey_OD,"../Results/pair-wise_prey_densiteis.png")

#get letters for multiple comparisons
multcompView::multcompLetters4(aov(mod_den_OD), pair_wise_OD)
```

## Visualise Density Data

1) plot mean densities of predator 

```{r , echo=FALSE}

p_mean_pred <-  pop_den_mean %>% 
  filter(!predator_history == "no predator") %>% 
  ggplot()+
  geom_jitter(aes(x = prey_history, y= mean_Pred, shape = interaction(prey_history,predator_history), colour = interaction(prey_history,replicate)), size = 2.5)+
  geom_crossbar(data =  filter(group_densities,!predator_history == "no predator"),aes(x = prey_history,y = avg_Pred, ymin= (-1*se_Pred)+avg_Pred, ymax = se_Pred+avg_Pred), colour = "#DC3220")+
  
  scale_shape_manual(breaks = c("ancestral prey.naive predator","evolved prey.naive predator","ancestral prey.evolved predator","evolved prey.evolved predator"),values = c(16,16,17,17), guide = "none")+
  scale_colour_manual(name = "replicate",breaks = c("ancestral prey.A","evolved prey.A","ancestral prey.C","evolved prey.C","ancestral prey.E","evolved prey.E"), values = c("#7cac00e6","#c87cffff","#7cac00e6", "#00bec4ff","#7cac00e6","#f8776eff"), guide = "none" )+
  
  
  ylab("predator density [cells/mL]")+
  xlab("prey evol. history")+
  facet_grid(~ predator_history )+
  theme_bw()+
  theme( plot.title = element_text(size = 20, face = "bold"),
         legend.text = element_text(size = 18),
         legend.title = element_text(size = 18, face = "bold"),
         legend.key.size = unit(1, "cm"),
         axis.text = element_text(color = "grey50", size = 12),
         axis.title = element_text(size = 12),
         axis.title.x = element_text(),
         axis.title.y = element_text( colour = "black"),
         axis.title.y.right = element_text( colour = "#00A9E0"),
         strip.text = element_text(size = 18, colour = "black", face = "bold"),
         strip.background = element_rect(color="black", fill="white", linewidth=1, linetype="solid"),
         panel.spacing = unit(.05, "lines"),
         panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), 
         axis.line = element_line(colour = "black"))


```


2) plot mean OD of prey

```{r, echo=FALSE}
p_mean_prey <- ggplot( )+
  geom_jitter(data = pop_den_mean, aes(x = prey_history,y = mean_OD, shape = interaction(prey_history,predator_history), colour = interaction(prey_history,replicate)), size = 2.5)+
  geom_crossbar(data = group_densities,aes(x = prey_history,y = avg_OD, ymin= (-1*se_OD)+avg_OD, ymax = se_OD+avg_OD), colour = "#005AB5")+
  
  scale_shape_manual(breaks = c("ancestral prey.naive predator","evolved prey.naive predator","ancestral prey.evolved predator","evolved prey.evolved predator","ancestral prey.no predator","evolved prey.no predator"),values = c(16,16,17,17,9,9), guide = "none")+
  scale_colour_manual(name = "replicate",breaks = c("ancestral prey.A","evolved prey.A","ancestral prey.C","evolved prey.C","ancestral prey.E","evolved prey.E"), values = c("#7cac00e6","#c87cffff","#7cac00e6", "#00bec4ff","#7cac00e6","#f8776eff"), guide = "none" )+
  
  
  
  ylab("optical density")+
  xlab("prey evol. history")+
  facet_grid(~ predator_history )+
  theme_bw()+
  theme( plot.title = element_text(size = 20, face = "bold"),
         legend.text = element_text(size = 18),
         legend.title = element_text(size = 18, face = "bold"),
         legend.key.size = unit(1, "cm"),
         axis.text = element_text(color = "grey50", size = 12),
         axis.title = element_text(size = 12),
         axis.title.x = element_text(),
         axis.title.y = element_text( colour = "black"),
         axis.title.y.right = element_text( colour = "#00A9E0"),
         strip.text = element_text(size = 18, colour = "black", face = "bold"),
         strip.background = element_rect(color="black", fill="white", linewidth=1, linetype="solid"),
         panel.spacing = unit(.05, "lines"),
         panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), 
         axis.line = element_line(colour = "black"))



```

3) combine both plots 

```{r, fig.dim=c(10,14)}
p_mean_pred / p_mean_prey + plot_layout(width = c(2,3)) + plot_annotation( title = "Predator and Prey mean density")

```

