---
title: "Pre-processing raw data"
author: "Julius Hoffmann"
date: "2024-01-26"
output: html_document
---
## Aim
The data I received from the collaborators at University of Turku (Teppo Hiltunen, Shane Hogle) is pre-processed for later analysis.
Some data was not relevant to the findings presented in the manuscript "Temporal changes in the role of species sorting and evolution determine community dynamics" and these are excluded.
Further, I also merge the counts data from 16s rRNA sequecning with the corresponding metadata.
The created files are then saved to be used in further analyses.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Load Packages

```{r , message=FALSE}
library(tidyverse)

```

## Load data



```{r, message=FALSE}
traits_exp <- read.csv("../Data/experiment_clones_traits.tsv", sep = "\t")

metadata <- read.csv("../Data/metadata.tsv", sep = "\t")
sequencing_counts <- read.csv("../Data/sequencing_counts.tsv", sep = "\t")

predator_density_raw <- read_tsv("../Data/ciliate_cell_counts.tsv")
prey_density_raw <-read_tsv("../Data/Bacterial Densities.txt")

```

## Phenotypic data

1) select OD values of the relevant time point (i.e. after 96h)
```{r}
OD_96h <- filter(traits_exp, incubation_length_hours == 96)
```

2) Calculate the defence value for the OD values after 96h.
Also only select relevant information.
```{r}
gPred_gAlone <- log(OD_96h [,"clone_plus_ciliateOD"]/OD_96h [,"cloneOD"])
OD_96h  <- cbind(OD_96h , gPred_gAlone)
OD_96h  <- OD_96h [,c(4,5,6,7,8,10,15)]
```
3) Save the created file to a dedicated folder

```{r}
write.csv(OD_96h,"../derived data/Clone_OD_96h.txt", row.names = TRUE)

```

## 16s rRNA sequencing
1) Joining both sequencing counts with meta data
```{r}
species_frequencies <- left_join(sequencing_counts, metadata,relationship = "many-to-many")
```

2) Select only the data relevant for this experiment (i.e. transfer volume 1800µl)

```{r}

species_frequencies <- species_frequencies %>% 
  filter(transfer_volume == "1800uL")

```

3) Save the created file to the dedicated folder

```{r}
write.csv(species_frequencies,"../Data/sequencing_counts and metadata.csv", row.names = FALSE)

```

## Population Densities

```{r}
predator_density_filtered <- predator_density_raw %>% 
  filter(replicate %in% c("A","C","E") & transfer_volume == "1800uL")

prey_density_filtered <-  prey_density_raw %>% 
  filter(replicate %in% c("A", "C", "E") & transfer_category == "hi") 
```

```{r}

write_delim(predator_density_filtered,"../derived data/predator_densities.tsv")

write_delim(prey_density_filtered,"../derived data/prey_densities.tsv")

```

